---
title: "Building interactive maps using Leaflet in R"
author: "Kate Pyper"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}

# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

## This Session

By the end of this session you'll be able to build maps like this:

[ADD MAP EXAMPLE]

---

## This Session

During this section we will work through how to build a map in R using Leaflet (~ 1 hour) followed by a practice task for you to work on in small groups supported by the Geospatial Cross Team (~ 1 hour).

In the first hour we will cover:

- Reading in and manipulating shapefiles

- Adding Base maps

- Adding areas and point markers (with legends)

- Adding and toggling layers

Before joining this session you should have installed the geospatial packages on Posit, and also set up your .Rprofile to run the geospatial set up code.

---

class: center, middle, inverse

# Reading in and manipulating shapefiles

---

# What are shapefiles?

Shapefiles contain information on how locations look on a map. This is usually some set of coordinates that make up the boundaries of an area or set of areas.

The shapefile is actually multiple files *with the same name* that link together to form a full picture of what the area looks like. For R the key files *needed at minimum* to construct a spatial object are:

- `.shp` - this file contains the boundary definitions for a set of areas 

- `.dbf` - this file contains additional data relating to the same set of areas as in the `.shp` file

- `.shx` - this is an index file which allows searching of the `.shp` file

Shapefiles are available for a variety of area types in Posit: 

```{r}

geo_dir <- "/conf/linkage/output/lookups/Unicode/Geography/"

``` 

These shapefiles contain optional files which provides richer spatial information (`.prj`) and faster loading.

---

# Reading in spatial data

The package that is recommended for all things spatial is `{sf}`. `{sf}` treats spatial data in R as a type of data frame called a simple features data frame - the simple features are the (not so simple) are definitions.

To read in a shapefile as a simple features data frame, use the command `read_sf()` and supply the name of the `.shp` file - r will automatically pull in the rest of the files.

To read in the locality level shape files:

```{r}

library(sf)
library(readr)
library(dplyr)

locality_shp <- read_sf(file.path(geo_dir, "Shapefiles/HSCP Locality (Datazone2011 Base)/HSCP_Locality.shp"))

```

This looks like a data frame, and can be treated in generally the same way as a data frame (with some additional quirks!)

- The column `geometry` contains the information on the area boundaries which comes from the `.shp` file

- The rest of the columns come from information in the `.dbf` file

---

# The Coordinate Reference System (CRS)

When you read in a shapefile it will have an associated coordinate system in place. you can check what this corrdinate system is:

```{r, out.lines = 4}
st_crs(locality_shp)
```

Our shapefile is on an ordnance survey grid - usually, we want to have a geography data in latitude longitude format. This corresponds to a coordinate reference system called EPSG:4326. To transform to this coordinate system use `st_transform()`:

```{r, out.lines = 4}
locality_shp <- st_transform(locality_shp, crs = 4326)

st_crs(locality_shp)
```

---

# Manipulating spatial data

In general you can use a simple features data frame in the same way as any other data frame - including using `{dplyr}` verbs like `filter()`, `mutate()` and `select()` (`select()` will automatically select the `geometry` column)

The main operation where you need to be a bit careful is when using joins, as these have an effect on the object type. The rule for these is that *the spatial object has to come first*.

When the spatial object comes first spatial attributes stay in place:

```{r, out.lines = 4}
locality_lookup <- readr::read_csv(file.path(geo_dir, "/HSCP Locality/HSCP Localities_DZ11_Lookup_20230804.csv")) 

left_join(locality_shp, locality_lookup, by = join_by(hscp_local == hscp_locality))
```

When the spatial object comes second they don't:

```{r, out.lines = 4}
left_join(locality_lookup, locality_shp, by = join_by(hscp_locality == hscp_local))
```
---
class: center, middle, inverse2

# Building a basic map

---

class: center, middle, inverse3

# Toggling layers

---