---
title: "Building interactive maps using Leaflet in R"
author: "Kate Pyper"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightSpans: true
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
library(sf)
library(readr)
library(dplyr)
library(leaflet)

geo_dir <- "/conf/linkage/output/lookups/Unicode/Geography/"

locality_shp <- read_sf(file.path(geo_dir, "Shapefiles/HSCP Locality (Datazone2011 Base)/HSCP_Locality.shp"))

locality_shp <- st_transform(locality_shp, crs = 4326)

gpprac <- read_csv("https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/54a6e1e3-98a3-4e78-be0d-1e6d6ebdde1d/download/practice_contactdetails_jan2024-open-data.csv") %>% 
  filter(HB == "S08000030") %>% 
  left_join({read_rds("/conf/linkage/output/lookups/Unicode/Geography/Scottish Postcode Directory/Scottish_Postcode_Directory_2024_1.rds") %>% select(Postcode = pc8, latitude, longitude)})

HSCP_shp <- read_sf(file.path(geo_dir, "Shapefiles/HSCP 2019/SG_NHS_IntegrationAuthority_2019.shp")) %>% 
  st_transform(crs = 4326)

list_rate <- gpprac %>%
  group_by(HSCP) %>% 
  summarise(Rate = 1000 * n()/sum(PracticeListSize)) %>% 
  right_join(HSCP_shp, ., by = join_by(HIACode == HSCP))


pal_clus <- colorFactor(scales::pal_dichromat("Categorical.12")(length(unique(gpprac$GPCluster))), gpprac$GPCluster)

pal_pop <- colorNumeric(phsstyles::phs_colors(c("phs-teal", "phs-magenta")), list_rate$Rate)
# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

## This Session

By the end of this session you'll be able to build maps like this:

```{r, echo = FALSE}
my_map <- leaflet(height = "75%") %>% 
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate),
              group = "Practices per 1,000 List Size") %>% 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}"),
                   group = "GP Cluster") %>% 
  addLegend(position = "bottomleft",
            title = "Practices per 1,000 List Size",
            pal = pal_pop,
            values = list_rate$Rate,
            group = "Practices per 1,000 List Size") %>% 
  addLegend(position = "bottomleft",
            title = "GP Cluster",
            pal = pal_clus,
            values = gpprac$GPCluster,
            group = "GP Cluster") %>% 
  addLayersControl(position = "topright",
                   options = layersControlOptions(collapsed = FALSE),
                   baseGroups = c("Practices per 1,000 List Size", "GP Cluster"))

my_map
```

---

## This Session

During this section we will work through how to build a map in R using Leaflet (~ 1 hour) followed by a practice task for you to work on in small groups supported by the Geospatial Cross Team (~ 1 hour).

In the first hour we will cover:

- Reading in and manipulating shapefiles

- Adding Base maps

- Adding areas and point markers (with legends)

- Adding and toggling layers

Before joining this session you should have installed the geospatial packages on Posit, and also set up your .Rprofile to run the geospatial set up code.

---

class: center, middle, inverse

# Reading in and manipulating shapefiles

---

# What are shapefiles?

Shapefiles contain information on how locations look on a map. This is usually some set of coordinates that make up the boundaries of an area or set of areas.

The shapefile is actually multiple files *with the same name* that link together to form a full picture of what the area looks like. For R the key files *needed at minimum* to construct a spatial object are:

- `.shp` - this file contains the boundary definitions for a set of areas 

- `.dbf` - this file contains additional data relating to the same set of areas as in the `.shp` file

- `.shx` - this is an index file which allows searching of the `.shp` file

Shapefiles are available for a variety of area types in Posit in the `Shapefiles` sub folder of: 

```{r}

geo_dir <- "/conf/linkage/output/lookups/Unicode/Geography/"

``` 

These shapefiles contain optional files which provides richer spatial information (`.prj`) and faster loading.

---

# Reading in spatial data

The package that is recommended for all things spatial is `{sf}`. `{sf}` treats spatial data in R as a type of data frame called a simple features data frame - the simple features are the (not so simple) definitions of how the data works as a map.

To read in a shapefile as a simple features data frame, use the command `read_sf()` and supply the name of the `.shp` file - r will automatically pull in the rest of the files.

To read in the locality level shape files:

```{r}

library(sf)
library(readr)
library(dplyr)

locality_shp <- read_sf(file.path(geo_dir, "Shapefiles/HSCP Locality (Datazone2011 Base)/HSCP_Locality.shp"))

```

This looks like a data frame, and can be treated in generally the same way as a data frame (with some additional quirks!)

- The column `geometry` contains the information on the area boundaries which comes from the `.shp` file

- The rest of the columns come from information in the `.dbf` file

---

# The Coordinate Reference System (CRS)

When you read in a shapefile it will have an associated coordinate system in place. you can check what this coordinate system is:

```{r, out.lines = 4}
st_crs(locality_shp)
```

Our shapefile is on an ordnance survey grid - usually, we want to have a geography data in latitude longitude format. This corresponds to a coordinate reference system called EPSG:4326. To transform to this coordinate system use `st_transform()`:

```{r, out.lines = 4}
locality_shp <- st_transform(locality_shp, crs = 4326)

st_crs(locality_shp)
```

---

# Manipulating spatial data

In general you can use a simple features data frame in the same way as any other data frame - including using `{dplyr}` verbs like `filter()`, `mutate()` and `select()` (`select()` will automatically select the `geometry` column)

The main operation where you need to be a bit careful is when using joins, as these have an effect on the object type. The rule for these is that *the spatial object has to come first*.

When the spatial object comes first spatial attributes stay in place:

```{r, out.lines = 4}
locality_lookup <- readr::read_csv(file.path(geo_dir, "/HSCP Locality/HSCP Localities_DZ11_Lookup_20230804.csv")) 

left_join(locality_shp, locality_lookup, by = join_by(hscp_local == hscp_locality))
```

When the spatial object comes second they don't:

```{r, out.lines = 4}
left_join(locality_lookup, locality_shp, by = join_by(hscp_locality == hscp_local))
```
---
class: center, middle, inverse2

# Building a basic map

---

# Base maps

One of the benefits to using leaflet is being able to add spatial information on top of a map of the area you are looking at. This is called the base map, and there are a few you can choose from. To add the default base map:

```{r}
library(leaflet)

leaflet() %>% # set up map
  addTiles()

```

---

# Base maps

If you want to add a different base map ([you can view available base maps here](https://leaflet-extras.github.io/leaflet-providers/preview/)):

```{r}
leaflet() %>% 
  addProviderTiles("Esri.WorldImagery")
```
---

# Markers

To add markers to your plot all you need is a set of coordinates containing latitude and longitude. There are a few different types of markers that you can add:

- Circle markers - These are equivalent to adding points to a map, can vary size and colour, add labels and pop-ups.

- Markers - These are location markers. There is a bit of work involved in customising these. 

- Awesome Markers - These are location markers that can be customised easily.

These will be demonstrated using the GP practice information:

```{r echo = FALSE}
gpprac <- read_csv("https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/54a6e1e3-98a3-4e78-be0d-1e6d6ebdde1d/download/practice_contactdetails_jan2024-open-data.csv") %>% 
  filter(HB == "S08000030") %>% 
  left_join({read_rds("/conf/linkage/output/lookups/Unicode/Geography/Scottish Postcode Directory/Scottish_Postcode_Directory_2024_1.rds") %>% select(Postcode = pc8, latitude, longitude)})
```

```{r, out.lines = 4}
gpprac
```
---

# Circle Markers

```{r}
leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(data = gpprac,
                   radius = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                Cluster: {gpprac$GPCluster}<br>
                                List Size: {gpprac$PracticeListSize}"))
```

---

# Markers

```{r}
leaflet() %>% 
  addTiles() %>% 
  addMarkers(data = gpprac,
             popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                Cluster: {gpprac$GPCluster}<br>
                                List Size: {gpprac$PracticeListSize}"))
```

---

# Awesome Markers

```{r}
leaflet() %>% 
  addTiles() %>% 
  addAwesomeMarkers(data = gpprac,
                    icon = awesomeIcons(markerColor = "red",
                                        icon = "glyphicon-plus"),
                    popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                Cluster: {gpprac$GPCluster}<br>
                                List Size: {gpprac$PracticeListSize}"))
```


---

# Customizing colour based on data

```{r}
pal <- colorFactor(scales::pal_dichromat("Categorical.12")(length(unique(gpprac$GPCluster))), gpprac$GPCluster)

leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(data = gpprac,
                   color = pal(gpprac$GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}"))
```

---

# Areas

We can also add in areas to the map. These can be filled with a colour that we specify manually, or it can be filled using a colour palette. To the previous map showing GP practices, let's add HSCP boundaries coloured by the number of GP practices per 1,000 list size.

```{r, eval = FALSE}
HSCP_shp <- read_sf(file.path(geo_dir, "Shapefiles/HSCP 2019/SG_NHS_IntegrationAuthority_2019.shp")) %>% 
  st_transform(crs = 4326)

list_rate <- gpprac %>%
  group_by(HSCP) %>% 
  summarise(Rate = 1000 * n()/sum(PracticeListSize)) %>% 
  right_join(HSCP_shp, ., by = join_by(HIACode == HSCP))


pal_clus <- colorFactor(scales::pal_dichromat("Categorical.12")(length(unique(gpprac$GPCluster))), gpprac$GPCluster)

pal_pop <- colorNumeric(phsstyles::phs_colors(c("phs-teal", "phs-magenta")), list_rate$Rate)

leaflet() %>% 
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate)) %>% 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}"))
```
---

# Areas

```{r, echo = FALSE}
HSCP_shp <- read_sf(file.path(geo_dir, "Shapefiles/HSCP 2019/SG_NHS_IntegrationAuthority_2019.shp")) %>% 
  st_transform(crs = 4326)

list_rate <- gpprac %>%
  group_by(HSCP) %>% 
  summarise(Rate = 1000 * n()/sum(PracticeListSize)) %>% 
  right_join(HSCP_shp, ., by = join_by(HIACode == HSCP))


pal_clus <- colorFactor(scales::pal_dichromat("Categorical.12")(length(unique(gpprac$GPCluster))), gpprac$GPCluster)

pal_pop <- colorNumeric(phsstyles::phs_colors(c("phs-teal", "phs-magenta")), list_rate$Rate)

leaflet() %>% 
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate)) %>% 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}"))
```
---

# Legends

We have to manually add legends to leaflet plots. We do this using `addLegend()`

```{r, eval = FALSE}
leaflet(height = "75%") %>% # Altered height to make the legends fit
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate)) %>% 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}")) %>% 
  addLegend(position = "topright",
            title = "Practices per 1,000 List Size",
            pal = pal_pop,
            values = list_rate$Rate) %>% 
  addLegend(position = "bottomright",
            title = "GP Cluster",
            pal = pal_clus,
            values = gpprac$GPCluster)
```

---
# Legends

```{r, echo = FALSE}
leaflet(height = "75%") %>% 
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate)) %>% 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}")) %>% 
  addLegend(position = "topright",
            title = "Practices per 1,000 List Size",
            pal = pal_pop,
            values = list_rate$Rate) %>% 
  addLegend(position = "bottomright",
            title = "GP Cluster",
            pal = pal_clus,
            values = gpprac$GPCluster)
```
---

class: center, middle, inverse3

# Toggling layers

---

# Setting Groups

One of the great things about Leaflet is the ability to add multiple pieces of information to a single plot, with the option for users to turn off the pieces of information that they don't want to see. 

Each `add` function (`addMarkers()`, `addPolygons()`, `addLegend()` etc.) has an argument called `group` which allows us to collect everything that relates to a particular group of information.

We can then `addLayersControl()` to create a control that allows us to switch groups on and off within the map.

As an example we'll combine the polygon and point maps that we've created using the GP data.

---

# Adding controls

```{r, eval = FALSE}
leaflet() %>% 
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate),
              `group = "PracticesSize"`) %>% 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}"),
                   `group = "Clusters"`) %>% 
  addLegend(position = "bottomleft",
            title = "Practices per 1,000 List Size",
            pal = pal_pop,
            values = list_rate$Rate,
            `group = "PracticesSize"`) %>% 
  addLegend(position = "bottomleft",
            title = "GP Cluster",
            pal = pal_clus,
            values = gpprac$GPCluster,
            `group = "Clusters"`) %>%  
  addLayersControl(position = "topright",
                   options = layersControlOptions(collapsed = FALSE),
                   `baseGroups = c("PracticesSize", "Clusters")`)  
```
---

```{r, echo = FALSE}
leaflet(height = "75%") %>% 
  addTiles() %>% 
  # Always add polygons first otherwise you cover up the points
  addPolygons(data = list_rate,
              color = ~pal_pop(Rate),
              group = "Practice Size") %>% #<< 
  addCircleMarkers(data = gpprac,
                   color = ~pal_clus(GPCluster),
                   radius = 2, fillOpacity = 1, opacity = 1,
                   popup = glue::glue("Practice: {gpprac$GPPracticeName}<br>
                                        Cluster: {gpprac$GPCluster}<br>
                                        List Size: {gpprac$PracticeListSize}"),
                   group = "Clusters") %>% #<< 
  addLegend(position = "bottomleft",
            title = "Practices per 1,000 List Size",
            pal = pal_pop,
            values = list_rate$Rate,
            group = "Practice Size") %>% #<< 
  addLegend(position = "bottomleft",
            title = "GP Cluster",
            pal = pal_clus,
            values = gpprac$GPCluster,
            group = "Clusters") %>% #<< 
  addLayersControl(position = "topright",
                   options = layersControlOptions(collapsed = FALSE),
                   baseGroups = c("Practice Size", "Clusters")) 
```
